Neural Connectome of the Ctenophore Statocyst

### 2025-05-06 manuscript/Manuscript.qmd
ref

In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy. Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity.

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in Platynereis dumerilii larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands, while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits.

Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly. This departs fundamentally from reflex-arc models of early nervous systems and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes, indicating spatially distinct coordination dynamics. These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia. In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.

This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells. Such a structure suggests a loop-based coordination mechanism, where the central circuit adapts its timing based on feedback from its target. Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms.

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization—but these remain hypotheses.

@@

As Jagar et al. (2010) once described, this organ evokes the metaphor of a “ciliary brain.” The network we describe here similarly orchestrates output through dynamic, self-regulating coordination. These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.

>>>>

In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy. Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.

Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates. All observed output synapses were directed to balancer ciliary cells. Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system. These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense. Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity [@keijzer2013; @dewiljes2015; @Veraszto2017].

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013; @Moroz2014]. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].

Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly. This departs fundamentally from reflex-arc models of early nervous systems [@arendt2021; @jékely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.

@@

In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ. Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b]. Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024]. These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.

As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes (Figures 4E and F), indicating spatially distinct coordination dynamics. These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia. In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.

This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells. Such a structure suggests a loop-based coordination mechanism [@arshavsky2003; @selverston2010], where the central circuit adapts its timing based on feedback from its target. Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

@@

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here similarly orchestrates output through dynamic, self-regulating coordination. These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.
